---
title: "Sonali Joshi - Assignment 5"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(GenomicRanges)
library(SummarizedExperiment)
library(GenomicFeatures)
library(DESeq2)
library(ggplot2)
library(dplyr)
```

## Exploratory Analysis
The Counts data and Phenotype data were loaded into R for analysis with Bioconductor.
Rows with very low counts were filtered out to reduce the size of the counts data.
```{r GetData}
CountsData<-read.delim('FeatureCountsData.txt', header = TRUE)
CountsData = CountsData[rowMeans(CountsData[,2:7])>5,]
head(CountsData,5)
PhenoData<-read.delim('Sample_phenotypes.txt',header=TRUE)
sample_data <- DataFrame(PhenoData)
head(PhenoData)

```

##Create DESeqDataSet object

The package DESeq2 was used for further analysis. A column called ExpDesign was added to the Phenotype table to label the fetal and adult samples. This is used by DESeq2 package to specify the experimental design for analysis. 
Construct the DESeqDataSet object from the Counts data and the phenotype data

```{r }
#data_dds <- DESeqDataSetFromMatrix(CountsData[,2:7], PhenoData, ~ExpDesign)
data_dds <- DESeqDataSetFromMatrix(CountsData[,2:7], PhenoData, ~ExpDesign)
```

The DESeq2 package recommends the use of raw data without normalizing for sequencing depth as it accounts for library size differences internally.

## Visualize data

The boxplot on raw data indicates that data transformation is needed before PCA.
```{r boxplot}

boxplot(counts(data_dds))

```
## Data Transformation
DESeq2 offers two transformations for count data to stabilize variance. Transform the data using the VST - Variance Stabilizing Transformation and plot the data.

```{r Transform}
vdata_dds <- vst(data_dds, blind = FALSE)
boxplot(vst(counts(data_dds)), col="blue")

```


## PCA
Explore the data further by doing PCA on the data, to check if the age (fetal, adults) or sex have correlations with the principal components.

```{r plots}
plotPCA(vdata_dds, c("ExpDesign"))
plotPCA(vdata_dds, c("Sex"))
```

It is clear from the PCA that age (Fetal, Adult) explains the difference in the counts data as compared to the sex.

